DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-05-05 manuscript/manuscript.tex
updating pdf

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4540}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

\abstract{%
Date estimates for times of evolutionary divergences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.
Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
}

@@

Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database (McTavish et al., 2015), age data stored in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.

@@

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013). DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across source chronograms.
\item
  The source chronogram(s) with the most input taxon names.
\item
  Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
\end{enumerate}

\emph{\textbf{Choosing a topology.--}}
DateLife requires a tree topology to summarize age data upon.
We recommend that users provide a tree topology as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

\emph{\textbf{Dating the topology.--}}
Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
These markers are aligned with MUSCLE by default (Edgar, 2004), or MAFFT (Katoh, Asimenos, \& Toh, 2009).
This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.

Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}. By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.

For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb, Ackerly, \& Kembel, 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
BLADJ does not incorporate branch lengths even when they are present.
When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes.
\textgreater{} NOTE: do you know how it decides which one to drop if two calibrations conflict? The parent or the child? It would drop both. It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.

@@

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
(Stephen A. Smith \& O'Meara, 2012).
The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the secondary calibrations as priors on node ages.
\textgreater{} NOTE: is this true?? Yes

\emph{\textbf{Visualizing results.--}}
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).

@@

At the time of writing of this manuscript (May 06, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

\texttt{datelife} is free and open source and it can be used through its current website
\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
\texttt{datelife}'s R package stable version is available
for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J. A., Mozzherin, D., Rees, T., \ldots{} Enquist, B. J. (2013). {The taxonomic name resolution service: An online tool for automated standardization of plant names}. \emph{{BMC Bioinformatics}}, \emph{14}(1). \url{https://doi.org/10.1186/1471-2105-14-16}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
Criscuolo, A., Berry, V., Douzery, E. J. P., \& Gascuel, O. (2006). {SDM: A fast distance-based approach for (super)tree building in phylogenomics}. \emph{{Systematic Biology}}, \emph{55}(5), 740--755. \url{https://doi.org/10.1080/10635150600969872}

@@

\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
Sanchez-Reyes, L. L., \& O'Meara, B. (2022). {\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}. \emph{R Package Release V0.2.2}. Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}

\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
Sanchez-Reyes, L. L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M. (2022). {\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}. \emph{R Package Release V0.6.2}. Retrieved from \url{https://doi.org/10.5281/zenodo.593938}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4362}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

\abstract{%
Date estimates for times of evolutionary divergences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.
Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
}

@@

Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database (McTavish et al., 2015), age data stored in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.

@@

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (\textbf{boyle2013?}). DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across source chronograms.
\item
  The source chronogram(s) with the most input taxon names.
\item
  Last but not least, single summary chronograms resulting from summarizing age data from available source chronograms, as follows:
\end{enumerate}

\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon. We recommend that users provide a tree topology as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019). Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007), or combine topologies from source chronograms using a supertree approach.

\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT, Katoh, Asimenos, \& Toh, 2009). Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).

\emph{\textbf{Combining source chronograms.--}} To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

\emph{\textbf{Congruifying nodes.--}} Once with a chosen topology, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}. By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.

\emph{\textbf{Summarizing congruified ages.--}}
For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic statistics (mean, median, minimum and maximum ages). These single summary ages per taxon pair are then used as secondary calibrations to date the tree topology.\\

\emph{\textbf{Dating the tree topology.--}}
By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data. This minimizes age variance in the resulting chronogram (Campbell O. Webb, Ackerly, \& Kembel, 2008).
When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.

@@

Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb et al., 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).

\emph{\textbf{Visualizing results.--}}
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.

@@

At the time of writing of this manuscript (May 05, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

\texttt{datelife} is free and open source and it can be used through its current website
\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
\texttt{datelife}'s R package stable version is available
for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
Sanchez-Reyes, L. L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M. (2022). {datelife: Scientific Data on Time of Lineage Divergence for Your Taxa}. \emph{R Package Version 0.6.2}. Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
